---
title: "Kidney Donor and Recipient Risk Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kidney_risk_scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The transplantr package includes vectorised functions to calculate a number of different deceased donor kidney risk indices which are based on Cox models of transplant survival and can be used to audit or risk adjust transplant outcome data.

The models included in version 0.1.0 are:

* UK Kidney Donor Risk Index (2019 version): `ukkdri()`
* UK Kidney Donor Risk Index (2012 version): `watson_ukkdri()`
* US Kidney Donor Risk Index: `uskdri()`
* UK Kidney Recipient Risk Index `ukkrri()`

The US risk index uses donor creatinine as a parameter, and the `uskdri()` function in this package 
uses creatinine measured in mcmol/l as the default for all functions, but creatinine can be entered in 
mg/dl either by setting the optional `units` parameter to "SI" or by calling the wrapper function `uskdri_US()` instead. All of the functions in transplantr using creatinine or bilirubin use units 
in the same way - for mg/dl either use `units == "US"` or call the `..._US()` wrapper function.

To start, load the transplantr library and I would also recommend using dplyr to incorporate the results of these functions into your datasets:

```{r setup}
library(transplantr)
```

# UK Kidney Donor and Recipient Risk Indices (2019)

The current UK Kidney Donor and Recipient Risk Indices are based on transplant and recipient survival data from the UK National Transplant Database and have been incorporated into the new deceased donor kidney matching scheme launched in September 2019, which aims to match expected transplant and recipient survival (although a number of other factors are also used in the algorithm).

## UK Kidney Donor Risk Index

The `ukkdri()` function calculates the donor risk and uses the following parameters:

Parameter | Description | Values
----------|-------------|-------
age | donor age in years | numeric vector
height | donor height in cm | numeric vector
htn | donor history of hypertension | numeric vector of values of 1 (yes) or 0 (no)
sex | donor sex | character vector of values of "F" or "M"
cmv | whether donor is CMV seropositive | numeric vector of values of 1 (yes) or 0 (no)
gfr | donor eGFR in ml/min | numeric vector
hdays | length of donor hospital stay | numeric vector

Using vectors of 1 and 0 is more efficient than TRUE/FALSE or "Male"/"Female", but if your dataset 
uses these values instead, it is easy enough to convert them with the dplyr::mutate() verb:

```{r, eval = F}
my_data2 = my_data %>%
  mutate(Donor.Hypertension = if_else(Donor.Hypertension == TRUE, 1, 0),
         Donor.CMV = if_else(Donor.CMV == "Pos", 1, 0))
```

Generating the results is also easy to do with _dplyr_:

```{r, eval = F}
my_data3 = my_data2 %>%
  mutate(UKKDRI = ukkdri(age = Donor.Age, height = Donor.Height, htn = Donor.Hypertension,
                         sex = Donor.Sex, cmv = Donor.CMV, gfr = Donor.eGFR, 
                         hdays = Donor.Hospital_Stay))
```

Although it is a vectorised function, `ukkdri()` will also work with single values:

```{r}
ukkdri(age = 50, height = 170, htn = 1, sex = "F", cmv = 0, gfr = 90, hdays = 2)
```

## UK Recipient Risk Index

The `ukkrri()` function calculates the UK Kidney Recipient Risk Index using the following parameters:

Parameter | Description | Values
---|---|---
age | recipient age in years | numeric vector
dx | whether recipient on dialysis at time of listing | numeric vector of 1 (yes) and 0 (no)
wait | waiting time since start of dialysis in days | numeric vector
dm | whether recipient has diabetes | numeric vector of 1 (yes) and 0 (no)

The easiest way to generate the UKKRRI data column is again to use dplyr::mutate() verb:

```{r, eval = F}
my_data4 = my_data3 %>%
  mutate(UKKRRI = ukkrri(age = Recipient.Age, dx = Recipient.OnDialysis, 
                         wait = Recipient.Waittime, dm = Recipient.Diabetes))
```

# UK Kidney Donor Risk Index (2012)

The first version of the UK Kidney Donor Risk Index was published by Watson et al. in 2012 and based on Cox regression of kidney transplant survival from the UK National Transplant Registry. It is simpler than the US KDRI and slighly outperforms the US KDRI when applied to the UK transpant series. It can be calculated using the `watson_ukkdri()` function, which includes these variables:

Parameter | Description | Values
---|---|---
age | donor age in years | numeric vector
htn | whether donor has history of hypertension | numeric vector of 1 = yes, 0 = no
weight | donor weight in kg | numeric vector
hdays | length of donor hospital stay | numeric vector
adrenaline | whether donor treated with adrenaline infusion | numeric vector of 1 = yes, 0 = no

It can be called within a _dplyr_ pipe as above, or as a single calculation:


```{r}
watson_ukkdri(age = 40, htn = 0, weight = 75, hdays = 0, adrenaline = 0)
```


# US Kidney Donor Risk Index

The US Kidney Donor Risk Index is a longer established risk stratification tool developed from American registry data, and can be calculated using the `uskdri()` function. The donor's serum creatinine is one of the parameters used, and like the other functions in _transplantr_ the default unit is µmol/l but can be changed to mg/dl either by setting the `units` parameter to `"US` or by calling the `uskdri_US()` wrapper function. The risk index function uses these variables:

Parameter | Description | Values
---|---|---
age | donor age in years | numeric vector
height | donor height in cm | numeric vector
weight | donor weight in kg | numeric vector
eth | donor ethnicity | character string vector of "black" or "non-black"
htn | whether donor has history of hypertension | numeric vector of 1 = yes, 0 = no
dm | whether donor has diabetes mellitus | numeric vector of 1 = yes, 0 = no
cva | whether donor cause of death was CVA | numeric vector of 1 = yes, 0 = no
creat | donor creatinine | numeric vector
hcv | whether donor has hepatitis C seropositive | numeric vector of 1 = yes, 0 = no
dcd | whether donor is DCD | numeric vector of 1 = DCD, 0 = DBD
scaling | OPTN Scaling Factor | single number (defaults to 1)
units | creatinine units | single string of "SI" for µmol/l or "US" for mg/dl

The Scaling Factor is the median KDRI for kidney donors in the USA in the previous year, and is published each year on the [https://optn.transplant.hrsa.gov/resources/allocation-calculators/kdpi-calculator/](OPTN website), 
together with a table for converting KDRI to KDPI, which gives a percentile of risk. For 2019 the scaling factor is approximately 1.2507. This normalises the donor risk within each year, but I think it obscures inter-year comparisons. The `scaling` parameter defaults to 1 so can be left out when a non-scaled KDRI is desired.

As a vectorised function, this works well when calculating US KDRI for a large series using a _dplyr_ pipe with code like this:

```{r, eval = F}
my_data5 = my_data2 %>%
  mutate(USKDRI = uskdri(age = Donor.Age, height = Donor.Height, weight = Donor.Weight,
                         eth = Donor.Race, htn = Donor.Hypertension, dm = Donor.Diabetes,
                         cva = Donor.CVA, creat = Donor.Creatinine,
                         hcv = Donor.HepatitisC, dcd = Donor.Type, scaling = 1.250697, units = "US"))
```

The function can also be used for a single case:

```{r}
# with creatinine in µmol/l (units = "SI" can be omitted)
uskdri(age = 40, height = 170, weight = 80, eth = "non-black", htn = 0, dm = 0,
    cva = 0, creat = 120, hcv = 0, dcd = 0, scaling = 1.250697, units = "SI")

# with creatinine in mg/dl and omitting scaling factor
uskdri(age = 40, height = 170, weight = 80, eth = "non-black", htn = 0, dm = 0,
    cva = 0, creat = 1.4, hcv = 0, dcd = 0, units = "US")
```
