---
title: "Kidney Donor and Recipient Risk Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kidney_risk_scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Calculating kidney risk indices

The transplantr package includes vectorised functions to calculate a number of different deceased donor kidney risk indices which are based on Cox models of transplant survival and can be used to audit or risk adjust transplant outcome data.

The models included in version 0.1.0 are:

* UK Kidney Donor Risk Index (2019 version): `ukkdri()`
* UK Kidney Donor Risk Index (2012 version): `watson_ukkdri()`
* US Kidney Donor Risk Index: `uskdri()`
* UK Kidney Recipient Risk Index `ukkrri()`

The US risk index uses donor creatinine as a parameter, and the `uskdri()` function in this package 
uses creatinine measured in mcmol/l as the default for all functions, but creatinine can be entered in 
mg/dl either by setting the optional `units` parameter to "SI" or by calling the wrapper function `uskdri_US()` instead. All of the functions in transplantr using creatinine or bilirubin use units 
in the same way - for mg/dl either use `units == "US"` or call the `..._US()` wrapper function.

To start, load the transplantr library and I would also recommend using dplyr to incorporate the results of these functions into your datasets:

```{r setup}
library(transplantr)
library(dplyr)
```

## UK Kidney Donor and Recipient Risk Indices (2019)

The current UK Kidney Donor and Recipient Risk Indices are based on transplant and recipient survival data from the UK National Transplant Database and have been incorporated into the new deceased donor kidney matching scheme launched in September 2019, which aims to match expected transplant and recipient survival (although a number of other factors are also used in the algorithm).

### UK Kidney Donor Risk Index

The `ukkdri()` function calculates the donor risk and uses the following parameters:

Parameter | Description | Values
----------|-------------|-------
age | donor age in years | numeric vector
height | donor height in cm | numeric vector
htn | donor history of hypertension | numeric vector of values of 1 (yes) or 0 (no)
female | whether donor is female | numeric vector of values of 1 (yes) or 0 (no)
cmv | whether donor is CMV seropositive | numeric vector of values of 1 (yes) or 0 (no)
gfr | donor eGFR in ml/min | numeric vector
hdays | length of donor hospital stay | numeric vector

Using vectors of 1 and 0 is more efficient than TRUE/FALSE or "Male"/"Female", but if your dataset 
uses these values instead, it is easy enough to convert them with the dplyr::mutate() verb:

```{r, eval = F}
my_data2 = my_data %>%
  mutate(Sex = if_else(Sex == "F", 1, 0),
         Hypertension = if_else(Hypertension == TRUE, 1, 0),
         CMV = if_else(CMV == "Pos", 1, 0))
```

Generating the results is also easy to do with dplyr:

```{r, eval = F}
my_data3 = my_data2 %>%
  mutate(UKKDRI = ukkdri(age = Donor.Age, height = Donor.Height, htn = Donor.Hypertension,
                         female = Donor.Sex, cmv = Donor.CMV, gfr = Donor.eGFR, 
                         hdays = Donor.Hospital_Stay))
```

Although it is a vectorised function, `ukkdri()` will also work with single values:

```{r}
ukkdri(age = 50, height = 170, htn = 1, female = 1, cmv = 0, gfr = 90, hdays = 2)
```

### UK Recipient Risk Index

The `ukkrri()` function calculates the UK Kidney Recipient Risk Index using the following parameters:

Parameter | Description | Values
---|---|---
age | recipient age in years | numeric vector
dx | whether recipient on dialysis at time of listing | numeric vector of 1 (yes) and 0 (no)
wait | waiting time since start of dialysis in days | numeric vector
dm | whether recipient has diabetes | numeric vector of 1 (yes) and 0 (no)

The easiest way to generate the UKKRRI data column is again to use dplyr::mutate() verb:

```{r, eval = F}
mydata4 = mydata3 %>%
  mutate(UKKRRI = ukkrri(age = Recipient.Age, dx = Recipient.OnDialysis, 
                         wait = Recipient.Waittime, dm = Recipient.Diabetes))
```

